Functional Pearl: the Proof Search Monad

### 2015-10-12 paper/main.tex
Draft conclusion + formulating search with substitutions.

\newcommand{\fref}[1]{Figure~\ref{fig:#1}}
\newcommand{\R}{\ensuremath{\mathcal{R}}} % Rigid
\newcommand{\f}[1]{\ensuremath{\mathbf{#1}}} % Bold
\newcommand{\F}{\ensuremath{\mathcal{F}}} % Flexible
\newcommand{\V}{\ensuremath{\mathcal{V}}} % The prefix

@@

\begin{abstract}
  We present the proof search monad, a set of combinators that allows one to
  write a proof search engine in a style that resembles closely the inference
  rules. The user calls functions such as \li+premise+, \li+prove+ or
  \li+choice+; the library then takes care of generating a derivation tree,
  which serves either as a witness (success) or as an error message (failure).
  Proof search engines written in this style enjoy: first, a one-to-one
  correspondence between the implementation and the theoretical rules, which
  makes manual inspection easier; second, proof witnesses ``for free'', which
  makes a verified-validation approach easier too.
\end{abstract}

@@

We consider conjunctions of equalities of the form
$\V x_1, \ldots, \V x_n.\ \bigwedge_k x_i = x_j$, that is, conjunctions of
equalities between variables. Some variables are \emph{flexible}, meaning they may be
substituted for other variables. Their binder $\V x_i$ is of the form $\F \f
x_i$ (and their are typeset in bold font). Other variables are \emph{rigid}, and
may not be substituted. Their binder $\V x_i$ is of the form $\R x_i$. The
problem consists in computing a substitution of \emph{flexible} variables that
evaluate the conjunction to \li+true+. If no such substitution exists, the
solver outputs nothing.

For instance, one may want to solve: $\R x, \R z, \F \f y.\ x = \f y \wedge z = z$.
A solution exists: the solver outputs $\sigma = \{ \f y \mapsto x \}$ as a valid
substitution that solves the input problem. However, if one attempts to solve:
$\R x, \R z, \F \f y.\ x = \f y \wedge \f y = z$, the solver fails to find a
proper substitution, and return nothing. Indeed, the first clause demands that
$\f y$ substitutes to $x$, meaning that the second clause becomes $x = z$, which
always evaluates to false ($x$ and $z$ are two distinct rigid variables).

@@

In order to simplify the discussion, we skip elimination rules and quantifiers
altogether, and just assume that our input problem contains a combination of
flexible and rigid variables. Furthermore, we assume that any instantiation of a
flexible variable is legal. The only crucial point is that the solver outputs
something.

(The library that we describe here has been used in the implementation of the
Mezzo type-checker. There, all introduction and elimination rules are present;
furthermore, the implementation uses \emph{levels} to guarantee that only sound
instantiation choices are performed. This is, however, orthogonal to the present
discussion; the minimal problem we consider already conveys all the important
points, so we will skip quantifiers in the remainder of the discussion.)

Solving conjunctions of equalities requires three rules, which make up our
logic (\fref{proof-system}). \Rule{Refl} embodies the reflexivity axiom; \Rule{And}
highlights that the proof produces an \emph{output} $V'$ which is \emph{chained}
from one premise to another; \Rule{Inst} embodies the ``instantiation''
mechanism: in essence, if $y$ has not been instantiated so far, one
may instantiate it onto any other variable and perform the substitution in order
to prove the goal ($[\f y/x] P$ stands for ``replace $y$ with $x$ in $P$'').

\begin{figure}
  \centering
  \begin{mathpar}
    \inferrule[Refl]{
      \quad
    }{
      V \vdash x = x \dashv V
    }

    \inferrule[Inst]{
      x \in V_i \\
      V_1, \F(\f y = x), V_2 \vdash [\f y/x] P \dashv V'
    }{
      V_1, \F(\f y), V_2 \vdash P \dashv V'
    }

    \inferrule[And]{
      V \vdash P \dashv V' \\
      V' \vdash Q \dashv V''
    }{
      V \vdash P \wedge Q \dashv V''
    }
  \end{mathpar}
  \caption{Proof rules for the first system}
  \label{fig:proof-system}
\end{figure}

Once the proof tree has been laid out, one obtains an output (e.g. $V''$ in
\Rule{And}) where the instantiation choices (e.g. $\F (\f x_i = x_j)$) determine
the desired substitution (e.g. $\sigma = \{ \f x_i \mapsto x_j \}^*$).

@@

The solver can now solve problems of the form $\R x, \R z, \F \f y.\ x = z \vee
\f y = z$. It fails, however, to solve problems of the form $\R x, \R z, \F \f
y.\ (\f y = x \vee \f y = z) \wedge \f y = z$. The reason is, the option monad
is not powerful enough: upon finding a suitable choice in the disjunction
case, it commits to it and drops the other one. In other words, when hitting the
disjunction, \li+MOption+ commits to $\sigma = \{ \f y \mapsto x \}$, instead of
keeping $\sigma = \{ \f y \mapsto z \}$ as a backup solution. Phrased yet again
differently, we need to replace \li+MOption+ with the non-determinism monad that
will implement \emph{backtracking}.

@@

Conceptually, we want to change our way of thinking; instead of thinking of
\li+solve+ as a function that returns \emph{a solution}, we now think of it as a
function that returns \emph{several possible solutions}. The state is now a
set of states, each of which represent a possible choice.

@@

In particular, the earlier example of $\R x, \R z, \F \f
y.\ (\f y = x \vee \f y = z) \wedge \f y = z$ is now successfully solved by the
library. Thanks to laziness, no extra computations occur; further solutions down
the lazy list are only evaluated if the first ones failed.

@@

>>>>

\newcommand{\fref}[1]{Figure~\ref{fig:#1}}
\newcommand{\R}{\ensuremath{\mathcal{R}}} % Rigid
\newcommand{\f}[1]{\ensuremath{#1^?}} % Bold
\newcommand{\F}{\ensuremath{\mathcal{F}}} % Flexible
\newcommand{\V}{\ensuremath{\mathcal{V}}} % The prefix

@@

\begin{abstract}
  We present the proof search monad, a set of combinators that allows one to
  write a proof search engine in a style that resembles closely the inference
  rules. The user calls functions such as \li+premise+, \li+prove+ or
  \li+choice+; the library then takes care of generating a derivation tree.
  Proof search engines written in this style enjoy: first, a one-to-one
  correspondence between the implementation and the theoretical rules, which
  makes manual inspection easier; second, proof witnesses ``for free'', which
  makes a verified, independent validation approach easier too.
\end{abstract}

@@

We consider conjunctions of equalities of the form
$\bigwedge_k x_i = x_j$, that is, conjunctions of
equalities between variables. Some variables are \emph{flexible}, meaning they may be
substituted with other variables. They are typeset as follows: $\f y$.
Rigid variables may not be substituted. They are typeset as follows: $x$. The
problem consists in computing a substitution of \emph{flexible} variables so that
the conjunction evaluates to \li+true+. If no such substitution exists, the
solver outputs nothing.

For instance, one may want to solve: $x = \f y \wedge z = z$.
A solution exists: the solver outputs $\sigma = \{ \f y \mapsto x \}$ as a valid
substitution that solves the input problem. However, if one attempts to solve:
$x = \f y \wedge \f y = z$, the solver fails to find a
proper substitution, and return nothing. Indeed, the first clause demands that
$\f y$ substitutes to $x$, meaning that the second clause becomes $x = z$, which
always evaluates to false ($x$ and $z$ are two distinct rigid variables).

@@

In order to simplify the discussion, we skip elimination rules and quantifiers
altogether, and just assume that our input problem contains a combination of
flexible and rigid variables. Furthermore, we assume that any instantiation of a
flexible variable is legal. The only point that matters is that the solver
computes a result which sequenced throughout the proof search.

(The library that we describe here has been used in the implementation of the
Mezzo type-checker. There, the logic (the type system) does have all four
introduction and elimination rules; in the implementation, we used \emph{levels}
to guarantee that only sound instantiation choices are performed. This is,
however, orthogonal to the present discussion; the minimal problem we consider
already conveys all the important points, so we skip quantifiers in the
remainder of the discussion.)

\fref{proof-system} describes a proof search procedure for conjunctions of
equalities. Rules poke at a set of variables $V$ to tell apart rigid variables from
flexible variables. Each rule takes a substitution $\sigma$ as an input, and outputs
an updated substitution $\sigma'$. A variable in $V$ is either a rigid ($x$),
or a flexible ($\f y$).

\Rule{Refl} embodies the reflexivity axiom; \Rule{And}
highlights that the first sub-goal (premise) produces an \emph{output} $V'$
which is then \emph{chained}
from one premise to another; \Rule{Inst} embodies the ``instantiation''
mechanism: in essence, if $\f y$ has not been instantiated so far, one
may instantiate it onto any other variable (rigid or flexible), i.e. add a new
entry for $\f y$ in the substitution. Finally, \Rule{Subst} may apply the
substitution at any time in order to prove the goal.

\begin{figure}
  \centering
  \begin{mathpar}
    \inferrule[Refl]{
      \quad
    }{
      V, \sigma \vdash x = x \dashv \sigma
    }

    \inferrule[Subst]{
      V, \sigma \vdash \sigma P \dashv \sigma'
    }{
      V, \sigma \vdash P \dashv \sigma'
    }

    \inferrule[Inst]{
      x \in V \\ \f y \in V \\ \f y \not\in \sigma \\\\
      V, \{ \f y \mapsto x \} \circ \sigma \vdash  P \dashv \sigma'
    }{
      V \vdash P \dashv \sigma'
    }

    \inferrule[And]{
      V, \sigma \vdash P \dashv \sigma' \\\\
      V, \sigma' \vdash Q \dashv \sigma''
    }{
      V \vdash P \wedge Q \dashv \sigma''
    }
  \end{mathpar}
  \caption{Semi-algorithmic proof rules}
  \label{fig:proof-system}
\end{figure}

Once the proof tree has been laid out, one obtains an output (e.g. $\sigma''$ in
\Rule{And}). Since flexible variables may instantiate onto other flexible
variables, one may need to apply the substitution several times. The reflexive-transitive
closure of $\sigma$, i.e. $\sigma^*$, is thus the desired result of the proof
search.

@@

The solver can now solve problems of the form $x = z \vee
\f y = z$. It fails, however, to solve problems of the form $(\f y = x \vee \f y = z) \wedge \f y = z$. The reason is, the option monad
is not powerful enough: upon finding a suitable choice in the disjunction
case, it commits to it and drops the other one. In other words, when hitting the
disjunction, \li+MOption+ commits to $\sigma = \{ \f y \mapsto x \}$, instead of
keeping $\sigma = \{ \f y \mapsto z \}$ as a backup solution. Phrased yet again
differently, we need to replace \li+MOption+ with the non-determinism monad that
will implement \emph{backtracking}.

@@

Conceptually, we want to change our way of thinking; instead of thinking of
\li+solve+ as a function that returns \emph{a solution}, we now think of it as a
function that returns \emph{several possible solutions}. The state is now a
set of states, each of which represent a path in the search tree of derivation
trees.

@@

In particular, the earlier example of $(\f y = x \vee \f y = z) \wedge \f y = z$ is now successfully solved by the
library. Thanks to laziness, no extra computations occur; further solutions down
the lazy list are only evaluated if the first ones failed.

@@

We presented a support library for writing a proof search engine using backtracking in
any given logic; indeed, the library is parameterized by: the type of formulas; the type of
rule applications; the internal state type of the client. By merely using the
combinators of the library, the client gets derivations built for free; this allows
a separate verifier to independently check the steps required to prove the
formula. By opting into the library, the client gets to rewrite their code in a
new syntactic style that makes rule application explicits, forbids ``bundled''
applications of multiple rules at the same time and clearly lays out the
premises required to prove a judgement. Since the code resembles the logical
rules, mistakes are easier to spot.

The logic presented in this paper is as simple as it gets. It does, however,
highlight the main concepts. A version of this library is used in the core of
Mezzo's type-checker. The version of the library used in Mezzo also builds
failed derivations; these failed derivations stop at the first failed premise
or, in case of a choice, list all the failed attempts. We have not yet explained
this last feature as a clean combination of monads and operators, but hope to do
so in the near future.
